---
title: "Covid-19 Drug Discovery Explorer"
author: "O'Donovan, Imami et al"
date: "2/24/2021"
output:
  html_document:
    includes:
      in_header: ganalytics.html
subtitle: 'Analysis Version: 1.10'
runtime: shiny
bibliography: references.bib
resource_files:
- ganalytics.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

library(dplyr)
library(readr)
library(ggplot2)
library(stringr)
library(shiny)
library(plotly)

dataset <- read_csv("https://raw.githubusercontent.com/AliSajid/Covid19/v1.10/results/ace2-summarized-dataset.csv") %>% 
  select(compound, avg, sdev)
```

## Introduction

This document is a companion document to the COVID-19 Drug Discovery paper by O'Donovan, Imami et al "Identification of new drug treatments to combat COVID19: A signature-based approach using iLINCS" et al published in *Scientific Reports*. The paper can be found on the journal site [*Identification of candidate repurposable drugs to combat COVID-19 using a signature-based approach*](https://www.nature.com/articles/s41598-021-84044-9) [@ODonovan2021]. This interactive document aims to visualize the dataset from our final list of candidate drugs. This version of the document utilizes the v1.10 of our [analysis repository](https://github.com/AliSajid/Covid19/tree/v1.10) (latest version available [here](https://github.com/AliSajid/Covid19)). The document sources the data from a list we have generated that is available at [this github link]("https://raw.githubusercontent.com/AliSajid/Covid19/v1.10/results/ace2-summarized-dataset.csv") as a Comma Separated Values (CSV) file.

## Summary

To summarize, in the paper we use a two-pronged transcriptomic approach to identify potential treatments for COVID-19, caused by SARS-CoV-2. 

We identified drugs currently in use (or being explored) as treatments for COVID-19 or other coronavirus-related illnesses, and clustered them according to their mechanisms of action and [ATC Classification](https://www.whocc.no/atc_ddd_index/). We then downloaded the drugs' iLINCS transcriptomic signatures, identified the 8 most common combinations of cell lines, times and concentrations, and averaged them to generate drug cluster signatures. 

We also downloaded the transcriptomic datasets generated in the Mt. Sinai SARS-CoV-2 study [GSE147507](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507) to generate a signature for SARS-CoV-2 infection. 

We then utilized the [iLINCS Portal](http://ilincs.org/ilincs/) to conduct connectivity analysis to identify candidate drug signatures that were both concordant with our drug cluster signatures and discordant with the SARS-CoV-2 signature (+/-0.321 threshold), resulting in the identification of `r nrow(dataset)` candidate drugs.

The list of candidate drugs was filtered further. Candidate drug signatures with a mean concordance $\geq0.47$ (standard deviation $\leq 0.08$) with signatures from all 8 cell line combinations that were used to generate drug cluster signatures, were selected. This resulted in `r nrow(dataset %>% filter(avg >= 0.47, sdev <= 0.08))` candidate drugs that were filtered further, based on FDA approval status, known antiviral efficacy and whether they are registered for clinical trial for the treatment of COVID-19.

## Analysis

This analysis is split into two parts. The [Visualization](#visualization) allows you to see the distribution of the drugs in  a scatter plot. The [Tabulation](#tabulation) part allows you to see the drugs and their associated values.

Visualization and Tabulation of candidate drugs values can be modulated using the control panel. In the control panel, you can set the the minimum mean value and the maximum standard deviation value for candidate drugs. It also allows you to filter out drugs that appear to be experimental (not FDA approved). The option to plot the candidate drugs that were filtered out is also available.

```{r control_panel}
inputPanel(
  sliderInput("thresh", label = "Minimum Value for Mean: ",
              min = 0.3, max = 0.8, value = 0.47, step = 0.01),
  sliderInput("sdev", label = "Maximum Value for Std. Deviation: ",
              min = 0, max = 0.15, value = 0.08, step = 0.01),
  radioButtons("approved", "Filter Drugs?", choices = c("Yes" = TRUE, "No" = FALSE), selected = TRUE),
  radioButtons("keep", "Plot Filtered Drugs?", choices = c("Yes" = TRUE, "No" = FALSE), selected = TRUE)
)

filtered_data <- reactive({
  
  if (input$approved) {
    fda <- dataset %>% 
    filter(str_detect(compound, "CHEMBL", negate = T),
         str_detect(compound, "SCHEMBL", negate = T),
         str_detect(compound, "^\\d+", negate = T),
         str_detect(compound, "^[A-Z]\\d*\\w*\\-?\\s?\\d+", negate = T),
         str_detect(compound, "[Ii]nhibitor", negate = T),
         str_detect(compound, "^Broad", negate = T),
         str_detect(compound, "^BRD*", negate = T),
         str_detect(compound, "^UNII", negate = T),
         str_detect(compound, "omer", negate = T),
         str_detect(compound, "^Tyrphostin", negate = T)
        )
  } else {
    fda <- dataset
  }
  
  if (input$keep) {
    final <- fda %>% 
      mutate(keep = if_else(avg >= input$thresh & sdev <= input$sdev, TRUE, FALSE))
  } else {
  final <- fda  %>% 
    filter(avg >= input$thresh, sdev <= input$sdev) %>% 
      mutate(keep = TRUE)
  }
  
  final %>% 
    mutate(
      text = str_glue("Drug: {compound}\nMean Concordance: {round(avg,4)}\nStd. Deviation: {round(sdev, 4)}")
    )
})

```


### Visualization {#visualization}

The following figure allows us to visualize the dataset. The x-axis has the mean concordance value and the y-axis has the standard deviation of the same. Other controls were described above.

The *Threshold* slider allows us to filter out the things that have a concordance score less than or equal to the value selected.

The *Std Deviation* slider allows us to filter out the things that have a standard deviation more than the selected value.

The FDA Approved filter allows us to (crudely) filter out putative FDA-unapproved-drugs. This filter is not stringent, to prevent exclusion of potential (approved) candidate drugs.

You can hover over individual points see their values.

```{r visualization}

renderText({
  if (input$approved) {
      str_glue(
        "With minimum concordance {input$thresh} and maximum deviation {input$sdev}, we have {nrow(filtered_data() %>% filter(keep == TRUE))} filtered drugs in our list."
    )
  } else {
      str_glue(
        "With minimum concordance {input$thresh} and maximum deviation {input$sdev}, we have {nrow(filtered_data() %>% filter(keep == TRUE))} drugs in our list."
      )
  }
})


renderPlotly({
  
  g <- ggplot(data = filtered_data(), mapping = aes(x = avg, y = sdev,
                                                    color = keep, text = text))
  
  
  label = str_glue("Minimum Mean: {input$thresh}\nMaximum SD: {input$sdev}\nDrugs Identified: {nrow(filtered_data() %>% filter(keep == TRUE))}")

  
p <- g + geom_hline(yintercept = input$sdev, color = "grey80", lwd = 2) +
    geom_vline(xintercept = input$thresh, color = "grey80", lwd = 2) +
    geom_point() + 
    xlab("Mean Concordance") + 
    ylab("Standard Deviation") +
    scale_x_continuous(breaks = seq(0.3, 0.8, 0.025), limits = c(0.4, 0.625)) +
    scale_y_reverse(breaks = seq(0, 0.15, 0.01), limits = c(0.10, 0)) +
    scale_color_manual(breaks = c(TRUE, FALSE), values = c("darkgreen", "lightgrey")) +
    guides(color = "none") +
    theme_minimal() +
    ggtitle("Concordance Plot of Identified Drugs") +
    theme(
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 12, face = "bold")
    ) +
    annotate("label", x = 0.6125, y = 0.01, label = label)

ggplotly(p, tooltip = "text") %>%
  layout(
    showlegend = FALSE,
    yaxis2 = list(
      scaleanchor = "x",
      scaleratio = 0.2,
      domain = c(0.55, 1)
    ),
    dragmode = "zoom"
  ) %>%
  add_annotations(
    x = 0.6125,
    y = -0.01,
    xref = "x",
    yref = "y",
    text = label,
    xanchor = 'center',
    showarrow = F,
    bordercolor = "black",
    bgcolor = "grey80"
  ) %>%
  config(
    displayModeBar = "static",
    collaborate = F,
    displaylogo = FALSE,
    modeBarButtonsToRemove = list(
      "sendDataToCloud",
      "toImage",
      "hoverClosestCartesian",
      "hoverCompareCartesian",
      "select2d",
      "lasso2d",
      "zoomIn2d",
      "zoomOut2d",
      "toggleSpikelines",
      "pan2d"
    )
  )

})

```

### Tabulation {#tabulation}

The following interactive table allows us to sift through the entire list along with the numerical values. The table uses the same filtering criteria as the scatterplot above. For ease of interpretation, the values have been rounded to 4 significant figures.

The columns of the table signify the following:

1. **Name**: The name of the drug that we have identified
2. **Mean Concordance**: The mean concordance score averaged across all 8 cell line datasets
3. **Std. Deviation**: The standard deviation of the score across our 8 cell line datasets

```{r tabulation}

renderDataTable({
  filtered_data() %>% 
    filter(keep == TRUE) %>% 
    select(-keep, -text) %>% 
    rename(
      Name = compound,
      `Mean Concordance` = avg,
      `Std. Deviation` = sdev
    ) %>% 
    mutate(across(where(is.numeric), round, 4))
})

```

# References

