---
title: "Covid19 Drug Discovery Explorer"
author: "O'Donovan et al"
date: "10/4/2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(ggplot2)
library(stringr)

dataset <- read_csv("https://raw.githubusercontent.com/AliSajid/Covid19/v1.6/results/sars2-summarized-dataset.csv") %>% 
  select(compound, avg, sdev)
```

## Introduction

This document aims to help us narrow down the results that we would like to be interested in. The document sources the data from a list we have generated that is available at [this github link](https://raw.githubusercontent.com/AliSajid/Covid19/master/results/sars2-summarized-dataset.csv).

The above mentioned list summarizes the results of our analysis in a concise way. The list has 5 columns which are as follows:

1. **compound**: The drug that we have extracted from iLINCS
2. **avg**: The mean concordance score averaged across our 8 cell line datasets
3. **sdev**: The standard deviation of the score across our 8 cell line datasets
4. **slog10**: The negative log base 10 of the standard deviation
5. **slog2**: The negative log base 2 of the standard deviation

## Exploration

The following section allows us to filter and visualize the data. 

The Threshold slider allows us to filter out the things that have a concordance score less than or equal to the value selected.

The Std Deviation slider allows us to filter out the things that have a standard deviation more than the selected value.

The FDA Approved filter allows us to filter out (crudely) putative FDA-unapproved-drugs. This filter leans towards allowing rather than excludding so as to not exclude an approved drug.

```{r eruptions, echo=FALSE}
inputPanel(
  sliderInput("thresh", label = "Pass Threshold: ",
              min = 0.3, max = 0.8, value = 0.4, step = 0.01),
  sliderInput("sdev", label = "Pass Std Deviation: ",
              min = 0, max = 0.15, value = 0.15, step = 0.01),
  checkboxInput("approved", "FDA Approved Only?", TRUE),
  checkboxInput("keep", "Keep Filtered Drugs?", FALSE)
)

filtered_data <- reactive({
  
  if (input$approved) {
    fda <- dataset %>% 
    filter(str_detect(compound, "CHEMBL", negate = T),
         str_detect(compound, "SCHEMBL", negate = T),
         str_detect(compound, "^\\d+", negate = T),
         str_detect(compound, "^[A-Z]\\d*\\w*\\-?\\s?\\d+", negate = T),
         str_detect(compound, "[Ii]nhibitor", negate = T),
         str_detect(compound, "^Broad", negate = T),
         str_detect(compound, "^BRD*", negate = T),
         str_detect(compound, "^UNII", negate = T),
         str_detect(compound, "omer", negate = T),
         str_detect(compound, "^Tyrphostin", negate = T)
        )
  } else {
    fda <- dataset
  }
  
  if (input$keep) {
    fda %>% 
      mutate(keep = if_else(avg >= input$thresh & sdev <= input$sdev, TRUE, FALSE))
  } else {
  fda  %>% 
    filter(avg >= input$thresh, sdev <= input$sdev) %>% 
      mutate(keep = TRUE)
  }
})

renderText({
  if (input$approved) {
      glue(
        "With minimum concordance {input$thresh} and maximum deviation {input$sdev}, we have {nrow(filtered_data() %>% filter(keep == TRUE))} (presumptive) FDA Approved drugs in our list."
    )
  } else {
      glue(
        "With minimum concordance {input$thresh} and maximum deviation {input$sdev}, we have {nrow(filtered_data() %>% filter(keep == TRUE))} drugs (including non-FDA-Approved) in our list."
      )
  }
})


renderPlot({
  
  g <- ggplot(data = filtered_data(), mapping = aes(x = avg, y = sdev, color = keep))
  
  g + geom_point() + xlab("Average Concordance") + ylab("Standard Deviation") +
    scale_x_continuous(breaks = seq(0.3, 0.8, 0.025), limits = c(0.3, 0.8)) +
    scale_y_continuous(breaks = seq(0, 0.15, 0.01), limits = c(0, 0.15)) +
    scale_color_manual(breaks = c(TRUE, FALSE), values = c("darkgreen", "lightgrey")) +
    guides(color = "none") +
    theme_minimal()
})


renderDataTable({
  filtered_data() %>% filter(keep == TRUE) %>% select(-keep)
})

```
